---
title: "Group Project"
author: "Saahil Rasheed"
date: "2/1/2020"
output:
  word_document: default
  html_document: default
---

```{r, message=FALSE}
library(car)
library(rio)
library(stargazer)
library(ggplot2)
library(dplyr)
library(Hmisc)
library(psych)
library(fastDummies)
library(plm)
library(nlme)
library(lmtest)
library(sandwich)
library(robustbase)
```

###Socio Economic Geography and Education
## Pre-Processing

```{r, message=FALSE}
rm(list =ls())
setwd("C:/Users/saahi/Dropbox/SDM/Socio-Economic-Geography-and-Education")

d <- import("2013-2017_School_Math_Results_-_Genderclean.xlsx")

# dummy variables for grade
# I created a dummy for each grade but should omit 1 dummy var from model as the base case
ddum <- dummy_cols(d,select_columns="grade")

# dummy variables for district
# I created a dummy for each borough but should omit 1 dummy var from model as the base case
ddum$bronx <-0
ddum$brook <- 0
ddum$man <- 0
ddum$queen <- 0
ddum$staten <- 0

# Boroughs are made up of the following districts 1-6 are Manhattan, 7-12 are Bronx, 13-23 & 32 are Brooklyn, 2
# 4-30 are Queens, and 31 is Statten Island

# populate borough dummy columns values
for (i in 1:7631)
{
  if (ddum$district[i] < 7)
  {
    ddum$man[i] = 1
  } else if (ddum$district[i] > 6 & ddum$district[i] < 12 )
  {
    ddum$bronx[i] = 1
  }else if (ddum$district[i] > 12 & ddum$district[i] < 24)
  {
    ddum$brook[i] = 1
  }else if (ddum$district[i] > 23 & ddum$district[i] < 31)
  {
    ddum$queen[i] = 1
  }else if (ddum$district[i] == 31)
  {
    ddum$staten[i] = 1
  }else if (ddum$district[i] == 32)
  {
    ddum$brook[i] = 1
  }
}

# remove rows that just had a 2013 or a 2017 score. the 0 value for the missing score was causing severe outlier
# issues and wildly throwing off the normality of the residuals with the models. Still seems to be some normality issues
# but maybe can remove a view outliers still with outlierTest() and then see if that improves things.
dclean <-filter(ddum,scalescore17 != 0)
dclean <-filter(dclean, scalescore13 !=0)
```

## Models

```{r}
# model for score change b/w years and boroughs - staten island as base case
m1 <- lm(scalescore17~scalescore13 + bronx + brook + man + queen, data=dclean)
summary(m1)

# model for score change difference between boys and girls
m2 <- lm(scalescore17~scalescore13 + female, data=dclean)
summary(m2)

# model for score change difference between grades - 3rd grade is base case
m3 <- lm(scalescore17~scalescore13 + grade_4 + grade_5 + grade_6 + grade_7 + grade_8, data=dclean)
summary(m3)
plot(m3)
# model for grades and gender - 3rd grade is base case
m4 <- lm(scalescore17~scalescore13 + female + grade_4 + grade_5 + grade_6 + grade_7 + grade_8, data=dclean)
summary(m4)

# model for grades and borough and female
m5 <- lm(scalescore17~scalescore13 + female + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 +
           grade_8, data=dclean)
summary(m5)
plot(m5)

m5.1 <- lm(scalescore17~poly(scalescore13,3) + female + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 +
           grade_8, data=dclean)
summary(m5.1)
resettest(m5.1)
plot(m5.1)

```


```{r}


m11 <- lm(scalescore17~log(scalescore13) + bronx + brook + man + queen, data=dclean)
summary(m11)

m12 <- lm(scalescore17~scalescore13 + bronx + brook + man + queen + I(scalescore13*bronx) , data = dclean)
summary(m12)

m13 <- lm(scalescore17~scalescore13 + bronx + brook + man + queen + I(scalescore13*bronx) + I(scalescore13^2) , data=dclean)
summary(m13)

m14 <- lm(scalescore17~scalescore13 + bronx + brook + man + queen + I(scalescore13*bronx) + I(scalescore13^2) + female, data=dclean)
summary(m14)

m15 <- lm(scalescore17~scalescore13 + bronx + brook + man + queen + I(scalescore13*bronx) + I(scalescore13^2) + female + grade_4 + grade_5 + grade_6 + grade_7 + grade_8, data=dclean)
summary(m15)

m16 <- lm(scalescore17~scalescore13 + bronx + brook + man + queen + female + grade_4 + grade_5 + grade_6 + grade_7 + grade_8 + I(grade_5*scalescore13-297.57) + I((scalescore13-297.57)*bronx) + I(scalescore13^2), data=dclean)
summary(m16)

plot(m16)
plot(m16$fitted.values,m16$residuals)
```
INTERPRETATION:

1.	M1: m1 is the original model.
2.	M11: level-log model which resulted in a decreased Adj R-sq value. Hence, I decided to go ahead without level-log or log-log transforms.
3.	M12: Made an interaction between scalescore13 and Bronx and this turned out to be significant. Additionally, the Bronx dummy variable is also significant in this model. (best model imo)
4.	M13: Took a transform of scalescore13 to the power of 2 in addition to the previous best model. Transform turned out to be significant. (BEST)
Decide to use transform of scalescore13 to the power of 2 in the upcoming models for different scenarios.

```{r}
d2 <- dclean[!is.na(dclean$scalescore13), ]
mean(d2$scalescore13)

stargazer(m13, m15, m16, type="text", title="Results", align=TRUE)
```

```{r}
# model for score change difference between boys and girls
m2 <- lm(scalescore17~scalescore13 + female, data=dclean)
summary(m2)

m21 <- lm(scalescore17~ scalescore13 + female + I(scalescore13^2) , data = dclean)
summary(m21)

m22<- lm(scalescore17~ scalescore13 + female + I(scalescore13^2) + I(scalescore13*female) , data = dclean)
summary(m22)
```
INTERPRETATION:

1. m2 is the original model.
2. Included transform of scalescore13^2. Improved R-sq. (BEST)
3. Included interaction between female and scalescore13. Interaction insignificant.

```{r}
# model for score change difference between grades - 3rd grade is base case
m3 <- lm(scalescore17~scalescore13 + grade_4 + grade_5 + grade_6 + grade_7 + grade_8, data=dclean)
summary(m3)

m31<- lm(scalescore17~scalescore13 + grade_4 + grade_5 + grade_6 + grade_7 + grade_8 + I(grade_5*scalescore13),data = dclean)
summary(m31)

m32<-lm(scalescore17~scalescore13 + grade_4 + grade_5 + grade_6 + grade_7 + grade_8 + I(grade_5*scalescore13) + I(grade_7*scalescore13) , data = dclean)
summary(m32)

m33<- lm(scalescore17~scalescore13 + grade_4 + grade_5 + grade_6 + grade_7 + grade_8 + I(scalescore13^2) , data = dclean)
summary(m33)
```
INTERPRETATION:

1. m3 is the original model.
2. m31: Included interaction between grade 5 and scalescore13.
3. m32: Included interaction between grade 7 and scalescore13. Insignificant.
4. m33: Only included transform of scalescore13 to m1. Improved adj-Rsq value. (BEST)

These interactions don't cause a big improvement in the adj R-sq value.

```{r}
# model for grades and gender - 3rd grade is base case
m4 <- lm(scalescore17~scalescore13 + female + grade_4 + grade_5 + grade_6 + grade_7 + grade_8, data=dclean)
summary(m4)


m41<- lm(scalescore17~scalescore13 + female + grade_4 + grade_5 + grade_6 + grade_7 + grade_8 + I(female*grade_5),data=dclean)
summary(m41)

m42<- lm(scalescore17~scalescore13 + female + grade_4 + grade_5 + grade_6 + grade_7 + grade_8 + I(grade_5*scalescore13) + I(scalescore13^2) , data=dclean)
summary(m42)
```
INTERPRETATION:

1. m4 is the original model.
2. m41: Included interaction between female and grade_5. Not a very big improvement in the adj R-sq.
3. m42: Added transform of scalescore13 to the power of 2 to model m41. Best model yet and all covariates turn out to be significant. (BEST)

```{r}
# model for grades and borough
m5 <- lm(scalescore17~scalescore13 + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 +
           grade_8, data=dclean)
summary(m5)


m51<- lm(scalescore17~scalescore13 + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 +
           grade_8 + I(scalescore13^2) + I(scalescore13^3) + I(grade_8*bronx) , data=dclean)
summary(m51)

m52<- lm(scalescore17~scalescore13 + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 +
           grade_8 + I(scalescore13^2) + I(grade_7*scalescore13) , data = dclean)
summary(m52)

m53<- lm(scalescore17~scalescore13 + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 +
           grade_8 + I(scalescore13^2) + I(grade_8*bronx) , data=dclean)
summary(m53)

```
INTERPRETATION:

1. m5 is the original model. 
2. m51: Took two transforms of scalescore13 to the power of 2 and 3 each. Added interaction between grade 8 and bronx boroughs. Best adj R-sq value.
3. m52: Took transform of scalescore13 to the power of 2. Added interaction between grade7 and scalescore13 and this interaction turned out to be insignificant. (BEST)
I tried out a number of these interactions taking grade with different boroughs but they turned out to be insignificant.
4. m53: Took transform of scalescore13 to the power of 2. Added interaction between grade 8 and bronx (taking the idea from model m51). Second best adj r-sq value.



#I chose the best models from the analysis done above and tested formally for normality. These models seem to pass the test for Normality of residuals.
```{r}
plot(m13$fitted.values,m13$residuals)
plot(m21$fitted.values,m21$residuals)
plot(m33$fitted.values,m33$residuals)
plot(m42$fitted.values,m42$residuals)
plot(m52$fitted.values,m52$residuals)

# check for normality of residuals
qqnorm(m13$residuals)
qqline(m13$residuals, col="red")

qqnorm(m21$residuals)
qqline(m21$residuals, col="red")

qqnorm(m33$residuals)
qqline(m33$residuals,col="red")

qqnorm(m42$residuals)
qqline(m42$residuals, col="red")

qqnorm(m52$residuals)
qqline(m52$residuals, col="red")
```


## Assumption Check

```{r}
# model for grades and borough and female

m5.1 <- lm(scalescore17~poly(scalescore13,3) + female + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 + grade_8, data=dclean)
summary(m5.1)
```


MLR 1: Linearity

```{r}
# set mtest to desired model
mtest = m5.1

summary(mtest)

# check for linearity in betas
plot(mtest$fitted.values,mtest$residuals, main="Residuals vs Fitted for Math Test Model")
plot(mtest)
```

MLR 2: Random Sampling & Outliers
hat values show no points are exerting leverage

```{r}
#check for outliers
qqPlot(mtest, simulate=T, labels=row.names(dclean)) 
outlierTest(mtest)

dclean1 = dclean[-219,]  # remove outlier from outlierTest

# plot leverage 
lev = hat(model.matrix(mtest))
plot(lev, pch=19, main="After Removed Leverage Points")
abline(6*mean(lev),0,col="red",lwd=3)

pts <-subset(lev,lev>3*mean(lev))
rpts = unique(match(pts,lev))

# length of rpts which is row num of leverage points
x=66

i=0
for (i in 0:65)
{
  x=x-i
  dclean1 = dclean1[-rpts[x],]
  x=66
}

#re-run model with outliers and leverage points removed
m5.2 <- lm(scalescore17~poly(scalescore13,3) + female + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 +
           grade_8, data=dclean1)

mtest = m5.2
```

MLR 3: Zero Conditional Mean - Endogineity

```{r}
# Zero cond. mean
# based on this we reject the null hypothesis say that we have some functional form misspecifications in the model.
resettest(mtest)
```


MLR 4: No Perfect Multicolinearity

```{r}
#No perfect multicolinearity

df = dplyr::select_if(dclean,is.numeric)
x = 19

cormat = rcorr(as.matrix(df))     # df is dataframe with all numbers
lower = cormat$P
upper = cormat$r

upper = as.data.frame(upper)
upper = round(upper,2)
lower = as.data.frame(lower)
lower = round(lower,2)

cormatout = upper
x = ncol(df)
for (i in 1:x){ #column         # x is the number of variables in df
  for (j in 1+i:x){ #row
    cormatout[j,i]=lower[j,i]
  }
}
cormat = cormatout[-j,]      # usually have to remove 1 or more empty rows
rm(lower,upper,df,cormatout)
stargazer(cormat,type="text", summary=FALSE, align=TRUE, digits=2, 
          title="Correlation Values over p-values for Math Scores")

vif(mtest)
```

MLR 5: No Heteroskedasticity

```{r}
# test for heteroskedasticity
bartlett.test(list(mtest$residuals,mtest$fitted.values))
leveneTest(mtest$residuals, mtest$fitted.values, center=mean)
```

MLR 6: Normality of Residuals
```{r}
# Normality
qqnorm(mtest$residuals)
qqline(mtest$residuals, col="red")
```

```{r}
# use robust SE for model
# using hetero robust standard errors
coeftest(mtest,vcov=vcovHC(mtest))      # I think we should use this one.

# another way of running regression on robust SE for Heteroskedasticity

m5.3 <- lmrob(scalescore17~poly(scalescore13,3) + female + bronx + brook + man + queen + grade_4 + grade_5 + grade_6 + grade_7 + grade_8, data=dclean1)


S(m5.3)
summary(m5.3)

plot(m5.3)


qqnorm(m5.3$residuals)
qqline(m5.3$residuals, col="red")

```

#### Graphs

 Boroughs are made up of the following districts 1-6 are Manhattan, 7-12 are Bronx, 13-23 & 32 are Brooklyn, 2
 4-30 are Queens, and 31 is Statten Island
```{r}
for(i in 1:nrow(dclean)){
  if(dclean$district[i]<7){
    dclean$borough[i]='Manhattan'
  }else if (dclean$district[i] > 6 & dclean$district[i] <= 12 )
  {
    dclean$borough[i] = 'Bronx'
  }else if (dclean$district[i] > 12 & dclean$district[i] < 24)
  {
    dclean$borough[i] = 'Brooklyn'
  }else if (dclean$district[i] > 23 & dclean$district[i] < 31)
  {
    dclean$borough[i] = 'Queens'
  }else if (dclean$district[i] == 31)
  {
    dclean$borough[i] = 'Staten Island'
  }else if (dclean$district[i] == 32)
  {
    dclean$borough[i] = 'Brooklyn'
  }
}

unique(dclean$borough)
```

Data visualization:
```{r}
library(lattice)


ggplot(dclean, aes(x=scalescore13, y=scalescore17, color = borough),size=1) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth(method=lm) + facet_grid(.~borough)+ ggtitle("Math Score by Boroughs in NYC")
  
```

```{r}
library(reshape2)
nine.new <- melt(dclean,id.vars='borough', measure.vars=c('scalescore13','scalescore17'))

nine.new <- nine.new[complete.cases(nine.new), ]
 
ggplot(data=nine.new,
       aes(x=borough, y=value, colour=variable)) +
       geom_line(size = 2)

```